Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2025-03-24 paper/paper.tex
spelling

In this section, we describe how to construct an intrinsically
verified parser for regular expressions by compiling it to an NFA and
then a DFA. That is, for each regular expression $A$, we construct an
NFA $N(A)$ and a corresponding DFA $D(A)$ such that $A$ is strongly
equivalent to the traces of $N(A)$ and weakly equivalent to the
accepting traces of $D(A)$. Then we can easily construct a parser for
traces of $D(A)$ and combine this using \Cref{lem:wk-eqv-parse} to get
a verified regex parser.

@@

In \cref{fig:nfatrace} we define a linear type of traces through an
NFA $\N$. $\Trace_{\N}$ is an inductive type indexed by the starting
state of the trace $\s : \N.\states$. Traces in $\N$ may be built
through one of three constructors. We may terminate a trace at an
accepting state with the constructor $\nil$. Here we use an Agda-style
unicode syntax for $\bigamp$, as well as using the function arrow to
mean a non-dependent version of $\bigamp$. If we had a trace beginning
at the destination state of a transition, then we may use the $\cons$
constructor to linearly combine that trace with a parse of the label
of the transition to build a trace beginning at the source of the
transition.  Finally, if we had a trace beginning at the destination
of an $\epsilon$-transition then we may use $\epscons$ to pull it back
along the $\epsilon$-transition and construct a trace beginning at the
source of the $\epsilon$-transition. As a shorthand, write $Parse_{N}$
for the accepting traces out of $N.init$.
\newcommand{\D}{\texttt{D}}
\begin{figure}
\begin{lstlisting}
data Trace(*@$_{\N}$@*) : (s : N.states) (*@$\to$@*) L where
  nil : (*@$\uparrow$@*)(&[ s : N.states ] N.isAcc s (*@$\to$@*) Trace(*@$_{\N}$@*) s)
  cons : (*@$\uparrow$@*)(&[ t : N.transitions ]
    ('N.label t' (*@$\lto$@*) Trace(*@$_{\N}$@*) (N.dst t) (*@$\lto$@*) Trace(*@$_{\N}$@*) (N.src t)))
  (*@$\epsilon$@*)cons : (*@$\uparrow$@*)(&[ t : N.(*@$\epsilon$@*)transitions ]
    (Trace(*@$_{\N}$@*) (N.(*@$\epsilon$@*)dst t) (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.(*@$\epsilon$@*)src t)))

@@

%% The first two equalities are by defintion. The third equality holds because
%% $\literal c$ is unambiguous and
%% $let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a$ and $c$ have type
%% $\literal c$. The final equality holds because our inductive hypothesis states
%% that $nil~1 \equiv tr$. Thus, each $A s$ and $Trace_{N_{c}}~s~true$ are strongly
%% equivalent --- in particular, $\literal c$ and $Parse_{N_{c}}$.

@@

\begin{corollary}[\Agda{\fillerlink}]
  We may build a parser for every regular expression $\R$.
\end{corollary}
\begin{proof}
We combine the strong equivalence of \cref{cons:thompson} with the weak
equivalance of \cref{cons:determinization} to show that $\R$ is weakly
equivalent to the traces of its determinized automaton. Then, we use
\cref{lem:wk-eqv-parse} to extend the parser from \cref{thm:dfa-parser}
with respect to this weak equivalence.
\end{proof}

@@

Our final example is of a simple grammar of arithmetic expressions
with an associative operation. Here we take the alphabet to be $\{
\stringquote{(}, \stringquote{)}, \stringquote{+} , \stringquote{NUM}
\}$. In \Cref{fig:binop-inductive} we define the expression grammar
using two mutually
recursive types, corresponding to the two non-terminals we would use
in a CFG syntax. The syntactic structure encodes that the binary
operation is right associative. In the same figure, we define the
traces of an automaton with one token of lookahead. The automaton\footnote{In
  the automaton definition,
  $\texttt{NotStartsWithLP}$ is defined as
  \(I \oplus
  (\literal{)} \oplus \literal{+} \oplus \mathtt{NUM}) \otimes \top
  \). Similarly,
  $\texttt{NotStartsWithRP}$ is defined as
  \(I \oplus
  (\literal{(} \oplus \literal{+} \oplus \mathtt{NUM}) \otimes \top
  \).
} has
four different ``states'', each with access to a natural number
``stack''. The ``opening'' state $O$ expects either an left paren, in
which case it increments the stack and stays in the opening state, or
sees a number and proceeds to the $D$ state. The ``done opening''
state $D$ is where lookahead is used: if the next token will be a
right paren, then we proceed to $C$; otherwise, we proceed to $A$.
In the ``closing'' state $C$ if
we observe a right paren, then we decrement the count and continue to the
$D$ state. In the ``adding'' state $A$, we succeed if the string
ends and the count is $0$; otherwise, if we see a plus we continue to the $O$
state. Additionally, since the automaton need parse all of the
incorrect strings, we add all of the failing cases.

@@

We also aim to adapt the approach used for our context-free grammar parserss for
$\LLL$ and $\LRR$/$\LALRR$ with the aim of developing a shared library of intrinsically
verified parsing utilities that can be usefully shared across different
parsers. We would also like to investigate how high-performance verified parsers
can be implemented in \theoryabbv. This might be done by developing an efficient
compiler for \theoryabbv directly or by developing a verified compiler to an
imperative system within \theoryabbv itself.

@@

\subsection{Grammar Semantics Respects Additional Axioms}
We verify that the denotational semantics verfies each of the axioms we have
assumed.

>>>>

In this section, we describe how to construct an intrinsically
verified parser for regular expressions by compiling it to an NFA and
then a DFA. That is, for each regular expression $A$, we construct an
NFA $N(A)$ and a corresponding DFA $D(A)$ such that $A$ is strongly
equivalent to the traces of $N(A)$ and weakly equivalent to the
accepting traces of $D(A)$. Then we can easily construct a parser for
traces of $D(A)$ and combine this using \Cref{lem:wk-eqv-parse} to get
a verified regular expression parser.

@@

In \cref{fig:nfatrace} we define a linear type of traces through an
NFA $\N$. $\Trace_{\N}$ is an inductive type indexed by the starting
state of the trace $\s : \N.\states$. Traces in $\N$ may be built
through one of three constructors. We may terminate a trace at an
accepting state with the constructor $\nil$. Here we use an Agda-style
Unicode syntax for $\bigamp$, as well as using the function arrow to
mean a non-dependent version of $\bigamp$. If we had a trace beginning
at the destination state of a transition, then we may use the $\cons$
constructor to linearly combine that trace with a parse of the label
of the transition to build a trace beginning at the source of the
transition.  Finally, if we had a trace beginning at the destination
of an $\epsilon$-transition then we may use $\epscons$ to pull it back
along the $\epsilon$-transition and construct a trace beginning at the
source of the $\epsilon$-transition. As a shorthand, write $Parse_{N}$
for the accepting traces out of $N.init$.
\newcommand{\D}{\texttt{D}}
\begin{figure}
\begin{lstlisting}
data Trace(*@$_{\N}$@*) : (s : N.states) (*@$\to$@*) L where
  nil : (*@$\uparrow$@*)(&[ s : N.states ] N.isAcc s (*@$\to$@*) Trace(*@$_{\N}$@*) s)
  cons : (*@$\uparrow$@*)(&[ t : N.transitions ]
    ('N.label t' (*@$\lto$@*) Trace(*@$_{\N}$@*) (N.dst t) (*@$\lto$@*) Trace(*@$_{\N}$@*) (N.src t)))
  (*@$\epsilon$@*)cons : (*@$\uparrow$@*)(&[ t : N.(*@$\epsilon$@*)transitions ]
    (Trace(*@$_{\N}$@*) (N.(*@$\epsilon$@*)dst t) (*@$\lto$@*) Trace(*@$_{\N}$@*)(N.(*@$\epsilon$@*)src t)))

@@

%% The first two equalities are by definition. The third equality holds because
%% $\literal c$ is unambiguous and
%% $let~a~\otimes~()~=~c~\otimes~fromN_{c}~1~tr~in~a$ and $c$ have type
%% $\literal c$. The final equality holds because our inductive hypothesis states
%% that $nil~1 \equiv tr$. Thus, each $A s$ and $Trace_{N_{c}}~s~true$ are strongly
%% equivalent --- in particular, $\literal c$ and $Parse_{N_{c}}$.

@@

\begin{corollary}[\Agda{\fillerlink}]
  We may build a parser for every regular expression $\R$.
\end{corollary}
\begin{proof}
We combine the strong equivalence of \cref{cons:thompson} with the weak
equivalence of \cref{cons:determinization} to show that $\R$ is weakly
equivalent to the traces of its determinized automaton. Then, we use
\cref{lem:wk-eqv-parse} to extend the parser from \cref{thm:dfa-parser}
with respect to this weak equivalence.
\end{proof}

@@

Our final example is of a simple grammar of arithmetic expressions
with an associative operation. Here we take the alphabet to be $\{
\stringquote{(}, \stringquote{)}, \stringquote{+} , \stringquote{NUM}
\}$. In \Cref{fig:binop-inductive} we define the expression grammar
using two mutually
recursive types, corresponding to the two non-terminals we would use
in a CFG syntax. The syntactic structure encodes that the binary
operation is right associative. In the same figure, we define the
traces of an automaton with one token of lookahead. The automaton\footnote{In
  the automaton definition,
  $\texttt{NotStartsWithLP}$ is defined as
  \(I \oplus
  (\literal{)} \oplus \literal{+} \oplus \mathtt{NUM}) \otimes \top
  \). Similarly,
  $\texttt{NotStartsWithRP}$ is defined as
  \(I \oplus
  (\literal{(} \oplus \literal{+} \oplus \mathtt{NUM}) \otimes \top
  \).
} has
four different ``states'', each with access to a natural number
``stack''. The ``opening'' state $O$ expects either a left paren, in
which case it increments the stack and stays in the opening state, or
sees a number and proceeds to the $D$ state. The ``done opening''
state $D$ is where lookahead is used: if the next token will be a
right paren, then we proceed to $C$; otherwise, we proceed to $A$.
In the ``closing'' state $C$ if
we observe a right paren, then we decrement the count and continue to the
$D$ state. In the ``adding'' state $A$, we succeed if the string
ends and the count is $0$; otherwise, if we see a plus we continue to the $O$
state. Additionally, since the automaton need parse all of the
incorrect strings, we add all of the failing cases.

@@

We also aim to adapt the approach used for our context-free grammar parsers for
$\LLL$ and $\LRR$/$\LALRR$ with the aim of developing a shared library of intrinsically
verified parsing utilities that can be usefully shared across different
parsers. We would also like to investigate how high-performance verified parsers
can be implemented in \theoryabbv. This might be done by developing an efficient
compiler for \theoryabbv directly or by developing a verified compiler to an
imperative system within \theoryabbv itself.

@@

\subsection{Grammar Semantics Respects Additional Axioms}
We verify that the denotational semantics validates each of the axioms we have
assumed.
